---
title: 'Economic and population health effects of storm events in the USA'
output: 
  html_document:
  keep_md: true
---

# Economic and population health effects of storm events in the USA

## Synopsis

This report contains an analysis of storm data for the United States of America (collected by NOAA) between 1950 and 2011. The report presents the top five most important types of storms in the USA both in terms of population health and economic impact.

## Data processing

### Read data
```{r}
library('knitr')
rm(list=ls(all=TRUE))
setwd('~/Documents/repdata_2/')
stormdata <- read.csv(
  bzfile('~/Downloads/repdata-data-StormData.csv.bz2'), 
  as.is=TRUE
)
```


### Prepare data for analysing the impact on population health

Impact on population health is collected in the dataset for fatalities and injuries. The following R code aggregates these two types of population health impacts by storm type and merges them together into a data.frame for reporting results. The top five impacts for fatalities and injuries are stored in the R objects `{r} ph1` and `{r} ph2`.

```{r prepdata}
pophealth.fatalities <- aggregate(stormdata$FATALITIES, 
  by=list(stormdata$EVTYPE), 
  FUN=sum
)
pophealth.injuries <- aggregate(stormdata$INJURIES, 
  by=list(stormdata$EVTYPE), 
  FUN=sum
)

names(pophealth.fatalities) <- c('Storm', 'Fatalities')
names(pophealth.injuries) <- c('Storm', 'Injuries')

# Merge the data by event type and subset top by event types
ph <- merge(pophealth.fatalities, pophealth.injuries)
ph1 <- ph[order(-ph$Fatalities),][1:5,]
ph2 <- ph[order(-ph$Injuries),][1:5,]
```

### Prepare data for analysing economy impact of storms

Impact on the economy is collected in the dataset for property and crop damage. Before aggregating this damage by storm type, the data must be transformed into common units as in the following R code.

```{r prepdata2}
stormdata$pdmg_mult <- 1
stormdata$pdmg_mult[which(toupper(stormdata$PROPDMGEXP) == 'B')] <- 1000000000
stormdata$pdmg_mult[which(toupper(stormdata$PROPDMGEXP) == 'M')] <- 1000000
stormdata$pdmg_mult[which(toupper(stormdata$PROPDMGEXP) == 'K')] <- 1000
stormdata$crop_mult <- 1
stormdata$crop_mult[which(toupper(stormdata$CROPDMGEXP) == 'B')] <- 1000000000
stormdata$crop_mult[which(toupper(stormdata$CROPDMGEXP) == 'M')] <- 1000000
stormdata$crop_mult[which(toupper(stormdata$CROPDMGEXP) == 'K')] <- 1000
```

The following R code aggregates these two types of economy impacts by storm type and merges them together into a data.frame for reporting results. The top five impacts for fatalities and injuries are stored in the R objects `{r} ec1` and `{r} ec2`. A third R object `{r} ec3` contains the total economic impact of the two types of damage. 

```{r prepdata3}

economy.property <- aggregate(
  (stormdata$PROPDMG*stormdata$pdmg_mult)/1000000, 
  by=list(stormdata$EVTYPE), 
  FUN=sum
)
economy.crop <- aggregate(
  (stormdata$CROPDMG*stormdata$crop_mult)/1000000, 
  by=list(stormdata$EVTYPE), 
  FUN=sum
)
  
names(economy.property) <- c('Storm', 'PropertyDamage')
names(economy.crop) <- c('Storm', 'CropDamage')

# Merge the data by event type and subset top by event types
ec <- merge(economy.property, economy.crop)
ec$TotalDamage <- ec$PropertyDamage + ec$CropDamage
ec1 <- ec[order(-ec$PropertyDamage),][1:5,]
ec2 <- ec[order(-ec$CropDamage),][1:5,]
ec3 <- ec[order(-ec$TotalDamage),][1:5,]
```

## Results

### Storm events and population health

The top five storm events in terms of fatalities:

```{r printpophealthdata1, echo=TRUE}
print(ph1)
```

The top five storm events in terms of injuries:

```{r printpophealthdata2, echo=TRUE}
print(ph2)
```

Summary of storm impacts on population health. Both fatalities and injuries considered:

```{r, fig.height=4, echo=TRUE}
par(mfrow=c(1,2))
barplot(
  ph1$Fatalities, 
  names.arg=ph1$Storm, 
  xlab='Fatalities', 
  main='Top 5 Most Deadly Storm Events'
)
barplot(
  ph2$Injuries, 
  names.arg=ph1$Storm,
  xlab='Injuries',
  main='Top 5 Storm Types Causing Injury'
)
```

### Storm events and the economy

There are two main economic impacts of storms in the USA, namely property and crop damage. 

The top 5 storm types by property damage are:
```{r economyimpact1, echo=TRUE}
print(ec1)
```

The top 5 storm types in terms of crop damage are:
```{r economyimpact2, echo=TRUE}
print(ec2)
```

When combined together, the types of storm events that cause the most negative impact on the USA economy are:
```{r economyimpact3, echo=TRUE}
print(ec3)
```

```{r, echo=FALSE}
econ_impact_1 <- ec3$Storm[1]
econ_impact_2 <- ec3$Storm[2]
```

The type of storm that causes the most total damage to the economy in the USA is `r econ_impact_1`, followed by `r econ_impact_2`.